### 2013-10-15 Publications/2013-TMI/04-methods.tex
Still working on nomenclature

\section{Methods}
\label{sec:methods}
%
\subsection{Related work}
\label{sec:methods_background}

@@

The first joint ``morphing''-segmentation approach (it was not a proper
registration method) was proposed in \citep{bertalmio_morphing_2000}. The first
registration framework is presented by \citep{yezzi_variational_2001} where the
energy functional is defined simultaneously in the moving and target images with
an affine transformation supporting the coordinates mapping (``2 PDEs approach'').
\citep{vemuri_joint_2003} proposed the first atlas-based registration based on the 
level set framework and using only one PDE. \citep{unal_coupled_2005} extended the 
idea of \citep{bertalmio_morphing_2000,yezzi_variational_2001} for non-linear 
registration with a dense deformation field as mapping function. \citep{droske_mumfordshah_2009}
combining ideas from both branches and proposed the propagation of the 
deformation field from the contours to the whole image definition. Finally,
\citep{gorthi_active_2011} present a comprehensive generalization of the methodologies.

@@

\subsubsection{Deformation field}
\label{sec:deformation_field}
The transformation from ``T1-'' into ``\gls{dwi}-'' coordinate space is 
achieved through a dense deformation field $u(\vec{x})$, such that:
\begin{equation}
\vec{c}_i' = U\{\vec{c}_i\} = \vec{c}_i + u(\vec{c}_i),
\label{eq:nodes_tfm}
\end{equation}
where $U$ is defined in \eqref{eq:deformation_field}. Since the nodes of the anatomical 
surfaces might lay off-grid, it is required to derive $u(\vec{x})$ from a discrete 
set of parameters $\{\vec{u}_k\}_{k=1 \ldots K}$. Densification is achieved through 
a set of associated basis functions $\psi_k$:
%
\begin{equation}
u(\vec{x}) = \sum_k \psi_k(\vec{x}) \vec{u}_k
\label{eq:intp_kernel}
\end{equation}
%
In our implementation, $\psi_k$ is chosen to be a tensor-product B-Spline kernel
of degree 3 ($B_3$).
Then, introducing \eqref{eq:intp_kernel} into \eqref{eq:nodes_tfm} and replacing
$\psi$ by the actual kernel function, the transformation writes
%
\begin{equation}
c_i' = c_i + \sum_k \left[ \vec{u}_k \, \underset{x,y,x}{\prod} B_3(c_{i,d} - \vec{x}_{k,d}) \right].
\label{eq:transformation}
\end{equation}

Based on the current estimate of the distortion $u$, we can compute 
``expected samples'' within the shape prior projected into the \gls{dwi}.
Thus, we now estimate region descriptors of the \gls{dwi} features 
$\vec{f}'$ of the regions defined by the priors in \gls{dwi} space.

\subsubsection{Gradient descent}
To find the minimum of the energy functional \eqref{eq:energy},
we propose a gradient-descent approach with respect to the underlying 
deformation field through the following partial differential equation:
\begin{equation}
\frac{\partial u(\vec{x},t)}{\partial t} \propto - \frac{\partial E(u)}{\partial u},
\label{eq:general_gradient_descent}
\end{equation}
with $t$ being an artificial time parameter of the contour
evolution, included to express the iterative character of the approach.
Now, we can introduce \eqref{eq:energy} in
\eqref{eq:general_gradient_descent}:
\begin{align}
\frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
\frac{ \partial }{\partial \vec{u}_k} \Big\{
C + \underset{l}{\sum}
\underset{\vec{x} \in \Omega_l}{\sum} \mdist{f'}{l} \notag\\
&+ \underset{\vec{x} \in \Omega}{\sum} \boldsymbol{\alpha} \cdot u(\vec{x})^{\circ2}
+ \underset{\vec{x} \in \Omega}{\sum} \boldsymbol{\beta} \cdot (\nabla \cdot u(\vec{x}))^{\circ2}
\Big\}.
\label{eq:gradient_descent}
\end{align}

@@

Therefore, we can apply a discretized interpretation of \eqref{eq:shape_gradients}
to compute the data term in \eqref{eq:gradient_descent} as follows:
\begin{align}
\frac{\partial E_{data}(u)}{\partial \vec{u}_k} &=
\underset{l}{\sum} \frac{ \partial }{\partial u^d_k} \lbrace
 \underset{\vec{x} \in \Omega_l}{\sum}
{\mathcal{D}_l^2}^2 (\vec{f}') \rbrace \notag\\
&= \underset{l,m}{\sum} \underset{\vec{c}' \in \Gamma_{l,m}}{\sum}
\left[ \Delta^2_i (\vec{f}' ) - \Delta^2_j (\vec{f}')\right]
\left\langle \frac{\partial \vec{c}'}{\partial {u^a_k}'}, \vec{n_i}'\right\rangle,
\label{eq:gradient_wshape}
\end{align}
in this case, the formulation has been adapted to the non-binary case, $l,m$
being any pair of neighboring regions, and $\Gamma_{l,m}$ the contour separating
them such that $\vec{x}' = \vec{c}' \in\Gamma_{l,m} \iff \vec{x}\in \partial\Omega_i \cap \partial\Omega_j$
and $\vec{n_i}'$ is the unit inward normal to the contour at $c_i'$.

Finally, we can compute:
\begin{align}
\frac{\partial}{\partial {\vec{u}_k}'}{\vec{c}_i}' &= \frac{\partial}{\partial {u^a_k}'} 
\lbrace c_i + \sum_k \psi_k(\vec{c}_i) \vec{u}_k \rbrace
= \psi_k(\vec{c_i})\, \hat{\vec{e}}^a
\end{align}
where $\hat{\vec{e}}^a$ is the unit vector along direction $a$. Projecting
this gradient onto the surface normal,
$\left\langle \frac{\partial \vec{c}'}{\partial {u^a_k}'}, \vec{n_i}'\right\rangle
= \psi_k(\vec{c_i})\, \hat{\vec{n}}_i^a$, the
full gradient evolution equation \eqref{eq:gradient_descent} yields:
\begin{align}
\frac{\partial E(u)}{\partial \vec{u}_k} 
&= \underset{l,m}{\sum} \underset{\vec{c}' \in \Gamma_{l,m}}{\sum}
\left[ \Delta^2_i (\vec{f}' ) - \Delta^2_j (\vec{f}')\right]
\psi_k(\vec{c_i})\, \hat{\vec{n}}_i^a \notag\\
&2\, \underset{x,y,z}{\sum} a_d u^a_{k,d}
-2\, \underset{x,y,z}{\sum} b_d \nabla u^a_{k,d},
\label{eq:gradient_final}
\end{align}

\subsubsection{Semi-implicit Euler-forward optimization}
It is necessary to discretize $t$ in order to obtain a numerical
implementation of the equation \eqref{eq:gradient_final}:
\begin{align}
\frac{u^{t+1}-u^{t}}{\tau} 
&= - \underset{l,m}{\sum} \underset{\vec{c}' \in \Gamma_{l,m}}{\sum}
\left[ \Delta^2_i (\vec{f}' ) - \Delta^2_j (\vec{f}')\right]
\psi_k(\vec{c_i})\, \hat{\vec{n}}_i^a \notag\\
&-2\, \underset{x,y,z}{\sum} a_d (u^a_{k,d})^{t+1}
+2\, \underset{x,y,z}{\sum} b_d \nabla (u^a_{k,d})^{t+1}.
\end{align}

The associated Euler-Lagrange equation is found as:
\begin{equation}
(\tau^{-1} +2\, a_d - 2\, b_d \nabla )\,(u^a_{k,d})^{t+1} = 
\tau^{-1} (u^a_{k,d})^{t} - \frac{\partial E_{data}(u)}{\partial u^d_k},
\end{equation}
that is a linear system that we translate into Fourier domain,
to obtain the next deformation field $(u^a_{k,d})^{t+1}$:
\begin{multline}
(u^a_{k,d})^{t+1} = \\
 \mathcal{F}^{-1} \left\lbrace
\frac{\mathcal{F}\lbrace \tau^{-1} (u^a_{k,d})^{t} - \underset{l,m}{\sum} \underset{\vec{c}' \in \Gamma_{l,m}}{\sum}
\left[ \Delta^2_i (\vec{f}' ) - \Delta^2_j (\vec{f}')\right] \rbrace}
     {\mathcal{F}\lbrace (\tau^{-1} +2\,a_d)\mathcal{I} - 2\,b_d \Delta ) \rbrace}
     \right\rbrace
\end{multline}

Here, we rewrite the Laplacian as a linear combination of the identity and shift operators:
\begin{equation}
\Delta = \sum\limits_d \mathcal{S}_d^- + \mathcal{S}_d^+ - 2 \mathcal{I}
\end{equation}
where $\mathcal{S}_{d}^{\pm}$ stands for the forward ($+$) and backward ($-$) shift 
operator along coordinates axis $d$, of which the Fourier transform is found easily as
\begin{equation}
\mathcal{F}\{\mathcal{S}_{d}^{\pm}\} = e^{\pm i\omega_{d}},
\end{equation}
where $\omega_{d}$ is the normalized pulsation along $d$-direction. Accordingly, the 
Fourier transform of the discrete Laplacian is found as
\begin{equation}
\mathcal{F}\{\Delta\} = \sum\limits_d e^{-i\omega_d } + e^{i\omega_d } - 2 = \sum\limits_d \left( 2\cos(\omega_d) - 2 \right)
\end{equation}

>>>>

1\section{Methods}
\label{sec:methods}
%
\subsection{Related work}
\label{sec:methods_background}

@@

The first joint ``morphing''-segmentation approach (it was not a proper
registration method) was proposed in \citep{bertalmio_morphing_2000}. The first
registration framework is presented by \citep{yezzi_variational_2001} where the
energy functional is defined simultaneously in the moving and target images with
an affine transformation supporting the coordinates mapping (``2 \glspl*{pde} approach'').
\citep{vemuri_joint_2003} proposed the first atlas-based registration based on the 
level set framework and using only one \gls*{pde}. \citep{unal_coupled_2005} extended the 
idea of \citep{bertalmio_morphing_2000,yezzi_variational_2001} for non-linear 
registration with a dense deformation field as mapping function. \citep{droske_mumfordshah_2009}
combining ideas from both branches and proposed the propagation of the 
deformation field from the contours to the whole image definition. Finally,
\citep{gorthi_active_2011} present a comprehensive generalization of the methodologies.

@@

\subsubsection{Deformation field}
\label{sec:deformation_field}
The transformation from ``T1-'' into ``\gls{dwi}-'' coordinate space is 
achieved through a dense deformation field $u(\vec{x})$, such that:
\begin{equation}
\vec{c}_i' = U\{\vec{c}_i\} = \vec{c}_i + u(\vec{c}_i),
\label{eq:nodes_tfm}
\end{equation}
where $U$ is defined in \eqref{eq:deformation_field}. Since the nodes of the anatomical 
surfaces might lay off-grid, it is required to derive $u(\vec{x})$ from a discrete 
set of parameters $\{\vec{u}_k\}_{k=1 \ldots K}$. Densification is achieved through 
a set of associated basis functions $\psi_k$:
%
\begin{equation}
u(\vec{x}) = \sum_k \psi_k(\vec{x}) \vec{u}_k
\label{eq:intp_kernel}
\end{equation}
%
In our implementation, $\psi_k$ is chosen to be a tensor-product B-Spline kernel
of degree 3 ($B_3$).
Then, introducing \eqref{eq:intp_kernel} into \eqref{eq:nodes_tfm} and replacing
$\psi$ by the actual kernel function, the transformation writes
%
\begin{equation}
\vec{c}_i' = \vec{c}_i + \sum_k \left[ \vec{u}_k \, \underset{n}{\prod} B_3( (\vec{c}_i - \vec{x}_k) \cdot \hat{\mathbf{e}}_d ) \right].
\label{eq:transformation}
\end{equation}

Based on the current estimate of the distortion $\vec{u}$, we can compute 
``expected samples'' within the shape prior projected into the \gls{dwi}.
Thus, we now estimate region descriptors of the \gls{dwi} features 
$\vec{f}'$ of the regions defined by the priors in \gls{dwi} space.

\subsubsection{Gradient descent}
To find the minimum of the energy functional \eqref{eq:energy},
we propose a gradient-descent approach with respect to the underlying 
deformation field through the following \gls*{pde}:
\begin{equation}
\frac{\partial u(\vec{x},t)}{\partial t} \propto - \frac{\partial E(\vec{u})}{\partial \vec{u}},
\label{eq:general_gradient_descent}
\end{equation}
with $t$ being an artificial time parameter of the contour
evolution, included to express the iterative character of the approach.
Now, we can introduce \eqref{eq:energy} in
\eqref{eq:general_gradient_descent}:
\begin{align}
\frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
\frac{ \partial }{\partial \vec{u}_k} \Big\{
C + \underset{l}{\sum}
\underset{\vec{x} \in \Omega_l}{\sum} \mdist{f'}{l} \notag\\
&+ \underset{\vec{x} \in \Omega}{\sum} \boldsymbol{\alpha} \cdot u(\vec{x})^{\circ2}
+ \underset{\vec{x} \in \Omega}{\sum} \boldsymbol{\beta} \cdot (\nabla \cdot u(\vec{x}))^{\circ2}
\Big\}.
\label{eq:gradient_descent}
\end{align}

@@

Therefore, we can apply a discretized interpretation of \eqref{eq:shape_gradients}
to compute the data term in \eqref{eq:gradient_descent} as follows:
\begin{align}
\frac{\partial E_{data}(\vec{u})}{\partial \vec{u}_k} &=
\underset{l}{\sum} \frac{ \partial }{\partial \vec{u}_k} \left\{
 \underset{\vec{x} \in \Omega_l}{\sum} \mdist{f'}{l} \right\} \notag\\
&= \underset{l,m}{\sum} \underset{i}{\sum}
\left[ \mdist{f_i'}{l} - \mdist{f_i'}{m} \right]
\left\langle \frac{\partial \vec{c}_i'}{\partial \vec{u}_k}, \vec{n_i}'\right\rangle,
\label{eq:gradient_wshape}
\end{align}
in this case, the formulation has been adapted to the non-binary case, $l,m$
being any pair of neighboring regions, and $\Gamma_{l,m}$ the contour separating
them such that $\vec{x}' = \vec{c}' \in\Gamma_{l,m} \iff \vec{x}\in \partial\Omega_i \cap \partial\Omega_j$
and $\vec{n_i}'$ is the unit inward normal to the contour at $c_i'$.

Finally, we can compute:
\begin{align}
\frac{\partial \vec{c}_i'}{\partial \vec{u}_k} &= \frac{\partial}{\partial \vec{u}_k} 
\left\{ \vec{c}_i + \sum_k \psi_k(\vec{c}_i) \vec{u}_k \right\} 
= \psi_k(\vec{c_i})\, \hat{\vec{e}}
\end{align}
where $\hat{\vec{e}}$ is the coordinates system's unit vector.
Projecting this gradient onto the surface normal,
$\left\langle \frac{\partial}{\partial \vec{u}_k}{\vec{c}_i}', \vec{n_i}'\right\rangle
= \psi_k(\vec{c_i})\, \hat{\vec{n}}_i$, then the
full gradient evolution equation \eqref{eq:gradient_descent} yields:
\begin{align}
\frac{\partial E(\vec{u}_k)}{\partial \vec{u}_k} =
&- \underset{l,m}{\sum} \underset{i}{\sum}
\left[ \mdist{f_i'}{l} - \mdist{f_i'}{m} \right]
\psi_k(\vec{c}_i)\, \hat{\vec{n}}_i \notag\\
&+2\, \boldsymbol{\alpha} \vec{u}_k
-2\, \boldsymbol{\beta} \Delta \vec{u}_k,
\label{eq:gradient_final}
\end{align}

\subsubsection{Semi-implicit Euler-forward optimization}
It is necessary to discretize $t$ in order to obtain a numerical
implementation of the equation \eqref{eq:gradient_final}:
\begin{align}
\frac{\vec{u}_k^{t+1}-\vec{u}_k^t}{\tau} =
&- \underset{l,m}{\sum} \underset{i}{\sum}
\left[ \mdist{f_i'}{l} - \mdist{f_i'}{m}\right]
\psi_k(\vec{c}_i)\, \hat{\vec{n}}_i \notag\\
&+2\, \boldsymbol{\alpha} \vec{u}_k^{t+1}
-2\, \boldsymbol{\beta} \Delta \vec{u}_k^{t+1}.
\end{align}

The associated Euler-Lagrange equation is found as:
\begin{equation}
(\tau^{-1} +2\, \boldsymbol{\alpha} - 2\, \boldsymbol{\beta} \Delta )\, \vec{u}_k^{t+1} = 
\tau^{-1} \vec{u}_k^t - \frac{\partial}{\partial \vec{u}_k} E_{data}(\vec{u}_k^t),
\end{equation}
that is a linear system that we translate into Fourier domain,
to obtain the next deformation field $\vec{u}_k^{t+1}$:
\begin{multline}
\vec{u}_k^{t+1} = \\
 \mathcal{F}^{-1} \left\lbrace
\frac{\mathcal{F}\lbrace \tau^{-1}\vec{u}_{k}^{t} - \underset{l,m}{\sum} \underset{i}{\sum}
\left[ \mdist{f_i'}{l} - \mdist{f_i'}{m}\right] \rbrace}
     {\mathcal{F}\lbrace (\tau^{-1} +2\,\boldsymbol{\alpha})\mathcal{I} - 2\,\boldsymbol{\beta} \Delta ) \rbrace}
     \right\rbrace
\end{multline}

Here, we rewrite the Laplacian as a linear combination of the identity and shift operators:
\begin{equation}
\Delta = \sum\limits_d \mathcal{S}_d^- + \mathcal{S}_d^+ - 2 \mathcal{I}
\end{equation}
where $\mathcal{S}_{d}^{\pm}$ stands for the forward ($+$) and backward ($-$) shift 
operator along coordinates axis $d$, of which the Fourier transform is found easily as
\begin{equation}
\mathcal{F}\{\mathcal{S}_{d}^{\pm}\} = e^{\pm \vec{i}\omega_{d}},
\end{equation}
where $\omega_{d}$ is the normalized pulsation along $d$-direction. Accordingly, the 
Fourier transform of the discrete Laplacian is found as
\begin{equation}
\mathcal{F}\{\Delta\} = \sum\limits_d e^{-\vec{i}\omega_d } + e^{\vec{i}\omega_d } - 2 = \sum\limits_d \left( 2\cos(\omega_d) - 2 \right)
\end{equation}
